Ecological performance of detecting data fabrication with summary statistics

### 2017-05-17 submission/manuscript.Rnw
Update study 1, add rmarkdown

Any field of empirical inquiry is faced with cases of scientific misconduct at some point, either in the form of fabrication, falsification or plagiarism (FFP). Psychology was faced with Stapel; medical sciences were faced with Poldermans and Macchiarini; life sciences were faced with Voignet. These are just a few examples of cases in the last decade. Overall, an estimated 2\% of all scholars have admitted to falsifying or fabricating research results at least once \citep{fanelli2009} and this is likely to be an underestimate due to socially desirable responses. The detection rate is likely to be even lower; for example, only around a dozen cases are discovered in the United States and the Netherlands, despite covering several hundreds of thousands of researchers. At best, this amounts to a detection rate far below 1\% of those 2\% who admit to fabricating data --- the tip of a seemingly much larger iceberg.

@@

In the last decade, detecting image manipulation has become one of the few forms of detecting scientific misconduct other than plagiarism. The Journal of Cell Biology scans each submitted image for potential manipulation \citep{The_Journal_of_Cell_Biology2015-vh}, which greatly increases the risk of detecting (blatant) image manipulation. More recently, algorithms have been developed to automate the scanning of images for (subtle) manipulations \citep{Koppers2016}. These developments in detecting image manipulation have increased detection risk during the pre-publication and post-publication phase by improving detection mechanisms and increasing the understanding of how images might be manipulated. Moreover, their application also helps researchers systematically evaluate research articles to estimate the extent of the problem of image manipulation \citep[4\% of all papers are estimated to contain manipulated images;][]{Bik06072016}.

Statistical methods can provide one way to improve detection of data fabrication in empirical research. Humans are notoriously bad at understanding and estimating probabilities \citep[e.g.,][]{tversky1974, Tversky1971}, which could manifest itself in the fundamentally probabilistic data they try to fabricate. That researchers do not understand probabilistic processes also presents itself in the interpretation of genuine research data \citep{Hoekstra2006, Sijtsma2015-ts, Goodman_1999}. When data are fabricated, probabilistic principles are easily violated if these principles are forgotten at the univariate level, bivariate level, trivatiate level, or beyond \citep{Haldane1948-nm}. Based on such a theoretical framework, statistical methods that investigate whether the reported data are actually plausible under theoretically probabilistic processes can be used to detect potential data fabrication.

The application of such statistical methods to detect data fabrication has occurred in several cases in recent years and has potential for future application beyond a case-basis. For example, problems with papers by Fuji were highlighted with statistical methods \citep{Carlisle2012-yg,Carlisle2015-wm}, resulting in 183 retractions \citep{oransky2015}. In this case, baseline measures across randomized groups were examined for too little variation. Random assignment should introduce a certain degree of random error that is might be missed by a human fabricator, misestimating the probabilistic process that generates such error. Another two cases are those of Sanna and Smeesters, where fabricated data were also detected with statistics \citep{simonsohn2013}. These cases inspected the variance of variances (i.e., the second level, or meta, variance). Once again, too little variation was what revealed problems in these data. These methods, although developed in a case-setting, need not be limited to cases. The application of such methods can be (semi-)automated if data are available in a machine-readable format that one of the statistical methods can be applied to. An example of such a potential case for mass application of using statistics to detect (potential) data fabrication is in the ClinicalTrials.gov database, where baseline measures across randomized groups are readily available for download and subsequent analysis \citep{Hartgerink2015-bm}.

Nonetheless, considering the potential harm of applying statistical methods to flag potentially problematic results, it needs to be sorted out whether such methods have diagnosticity that actually makes it responsible to apply them. We hardly know how researchers might go about fabricating data. Cases such as Fuji, Smeesters, and Sanna provide some insights, but are highly pre-selected (i.e., those who got caught/confessed) and as such, systematically biased. Relatively extensive descriptions in rare and partial autobiogrophical accounts provide little insight into the actual data fabrication process, except for the setting where it might take place \citep[e.g., late at night when no one is around;][]{stapel_book}. Additionally, the performance of methods to detect data fabrication is highly dependent on the unknown  prevalence of data fabrication and the power to actually to detect data fabrication. Given that we do not know how researchers might fabricate data, the diagnosticity of these methods cannot realistically be simulated.

The effectiveness of detection mechanisms and their consequences, hence their expected deterrence, is exacerbated by the increased usage of public online discussion platforms such as PubPeer (\url{https://pubpeer.com}). PubPeer serves as an "online journal club" where anyone can discuss articles. Authors are notified when someone leaves a response, providing them with the possibility to respond. Such a platform allows for public discussion of the paper, including discussion of reanalyses, methodology, availability of materials, etc. When detection mechanisms are freely available to use, they can lead to (a surge of) comments when applied by users. Recently, in 2016, one user (the main author of this paper) used `statcheck` software to report potential statistical reporting inconsistencies for 50,000 psychology articles, which led to a large discussion about (automated) online comments \citep{Baker_2016}. The impact of such new possibilities is not to be underestimated, although its potential to contribute to the scientific discussion should also not be. Nonetheless, the `statcheck` software was well validated prior to this application \citep{Nuijten_2015} and the same principle applies to the application of statistical methods to detect (potential) data fabrication.

Throughout this paper, we inspect statistical methods to detect data fabrication that can be applied to (1) summary results or (2) raw data. Even though the data available can look different depending on the structure of a study, there are certain common characteristics of results and the underlying raw data that can be inspected. For example, summary results frequently include means, standard deviations, test-statistics, and $p$-values. Raw data frequently contain at least some variables measured at a interval- or ratio scale \citep{Stevens_1946}.

We present a set of studies that directly test the validity of statistical methods to detect data fabrication. To this end, Study 1 inspected the performance of statistical methods to detect data fabrication when using only summary results (e.g., means and standard deviations) as typically reported in empirical research articles. Study 2 inspected the performance of statistical methods aimed at detecting data fabrication in raw data (i.e., the data underlying summary results). These two studies provide a first indication of how applicable and effective statistical methods are to detect data fabrication in practice, with actual researchers fabricating actual data.

@@

Second, we applied the reversed Fisher method to detect data fabrication to the nonsignificant $p$-values twice: once for the results of gender effects hypothesis in each study and once for the results of the interaction effect hypothesis for each study. The Fisher method \citep{fisher1925} tests for evidence of an effect in a set of $p$-values by testing for a right-skew $p$-value distribution, but we adjusted it here to test for results that are overly consistent with the null hypothesis and result in a left-skew distribution (see Figure \ref{leftskew}). The original Fisher method is computed as
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
\label{fishertest}
\end{equation}
and tests for right-skew in a set of $p$-values, but we adjust it to the following
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
\label{fisheradjusted}
\end{equation}
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. We set this threshold to .05 in order to include only nonsignificant test results. The theoretical idea behind this method is that researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution, given the frequent misinterpretation of $p$-values \citep[e.g., as the probability of an effect,][]{Goodman2008135,Altman_1995}. % I can add more refs but it is just a rabbit hole to start with more.

<<figure, echo = FALSE, results = hide>>=
if(!file.exists(sprintf('%sfigures/fisherfig.pdf', x12)))
{
  pdf(sprintf('%sfigures/fisherfig.pdf', x12), width = 14, height = 7)
  par(mar = c(5, 5, 3, 1), mfrow = c(1, 2))

  plot(dbeta(seq(0, 1, .001), .05, 1),
       x = seq(0, 1, .001), type = 'l',
       bty = 'n', xlab = 'P-value', ylab = 'Density',
       main = '(A) Right-skew (normal)')
  lines(dunif(seq(0, 1, .001)),
        x = seq(0, 1, .001), lty = 2)
  plot(dbeta(seq(0, 1, .001), 1, .05),
       x = seq(0, 1, .001), type = 'l',
       bty = 'n', xlab = 'P-value', ylab = 'Density',
       main = '(B) Left-skew (potentially anomalous)')
  lines(dunif(seq(0, 1, .001)),
        x = seq(0, 1, .001), lty = 2)
  abline(v = .05, lty = 3)
  text(x = .15, y = 33, labels = "<- Threshold")
  dev.off()
}
@

% add plot of left and right skew
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../figures/fisherfig.pdf}
\caption{Conceptual representation of what the Fisher test inspects (Equation \ref{fishertest}; panel A) and what the adjusted Fisher test inspects (Equation \ref{fisheradjusted}; panel B). Both panels test whether there is sufficient evidence that the solid line deviates from the dashed line, except that the type of deviation that the test is sensitive to is the exact opposite.}
\label{leftskew}
\end{center}
\end{figure}

>>>>

Any field of empirical inquiry is faced with cases of scientific misconduct at some point, either in the form of fabrication, falsification or plagiarism (FFP). 
Psychology was faced with Stapel; medical sciencoes were faced with Poldermans and Macchiarini; life sciences were faced with Voignet. These are just a few examples of misconduct cases in the last decade. Overall, an estimated 2\% of all scholars have admitted to falsifying or fabricating research results at least once \citep{fanelli2009} and this is likely to be an underestimate due to socially desirable responses. The detection rate is likely to be even lower; for example, only around a dozen cases are discovered in the United States and the Netherlands, despite covering several hundreds of thousands of researchers. At best, this amounts to a detection rate far below 1\% of those 2\% who admit to fabricating data --- the tip of a seemingly much larger iceberg.

@@

In the last decade, detecting image manipulation has become one of the few forms of detecting scientific misconduct (alongside plagiarism).
The Journal of Cell Biology scans each submitted image for potential manipulation \citep{The_Journal_of_Cell_Biology2015-vh}, which greatly increases the risk of detecting (blatant) image manipulation. More recently, algorithms have been developed to automate the scanning of images for (subtle) manipulations \citep{Koppers2016}. These developments in detecting image manipulation have increased detection risk during the pre-publication and post-publication phase by improving detection mechanisms and increasing the understanding of how images might be manipulated. Moreover, their application also helps researchers systematically evaluate research articles to estimate the extent of the problem of image manipulation \citep[4\% of all papers are estimated to contain manipulated images;][]{Bik06072016}.

Statistical methods can provide one way to improve detection of data fabrication in empirical research. Humans are notoriously bad at understanding and estimating probabilities \citep[e.g.,][]{tversky1974, Tversky1971}, which could manifest itself in the fundamentally probabilistic data they try to fabricate. 
That researchers do not understand probabilistic processes also presents itself in false interpretation of genuine research data \citep{Hoekstra2006, Sijtsma2015-ts, Goodman_1999}. When data are fabricated, probabilistic principles are easily violated if forgotten at the univariate level, bivariate level, trivatiate level, or beyond \citep{Haldane1948-nm}. Based on this idea, statistical methods that investigate whether the reported data are feasible under the theoretically probabilistic processes can be used to detect potential data fabrication.

The application of such statistical methods to detect data fabrication has occurred in several cases in recent years and has potential for future application beyond a case-basis. 
For example, problems in papers by Fuji were highlighted with statistical methods \citep{Carlisle2012-yg,Carlisle2015-wm}, resulting in 183 retractions \citep{oransky2015}. In this case, baseline measures across randomized groups were examined for too little variation. Random assignment should introduce a certain degree of random error that might be missed by a human fabricator, misestimating the probabilistic process that generates such error. Another two cases are those of Sanna and Smeesters, where fabricated data were also detected with statistics \citep{simonsohn2013}. These cases inspected the variance of variances (i.e., the second, or meta level variance). Once again, too little variation was what revealed problems in these data \citep[other examples include ][]{lacour-complaint,foerster-complaint}. These methods, although developed in a case-setting, need not be limited to cases. The application of such methods can be (semi-)automated if data are available in a machine-readable format that one of the statistical methods can be applied to. An example of such a potential case for mass application of using statistics to detect (potential) data fabrication is in the ClinicalTrials.gov database, where baseline measures across randomized groups are readily available for download and subsequent analysis \citep{Hartgerink2015-bm}.

Nonetheless, considering the potential harm of applying statistical methods to flag potentially problematic results, it needs to be sorted out whether such methods have diagnosticity that actually makes it responsible to apply them. 
We hardly know how researchers might go about fabricating data. Cases such as Fuji, Smeesters, and Sanna provide some insights, but are highly pre-selected (i.e., those who got caught/confessed) and as such, systematically biased. Relatively extensive descriptions in rare and partial autobiogrophical accounts provide little insight into the actual data fabrication process, except for the setting where it might take place \citep[e.g., late at night when no one is around;][]{stapel_book}. Trustworthiness of these accounts can be called into question. Additionally, the performance of methods to detect data fabrication is highly dependent on the unknown prevalence of data fabrication and the power to actually to detect data fabrication. Given that we do not know how researchers might fabricate data, the diagnosticity of these methods cannot realistically be simulated.

The effectiveness of detection mechanisms and their consequences, hence their expected deterrence, is exacerbated in the digital age by the increased usage of public online discussion platforms such as PubPeer (\url{https://pubpeer.com}). 
PubPeer serves as an "online journal club" where anyone can discuss articles. Authors are notified when someone leaves a response, providing them with the possibility to respond. Such a platform allows for public discussion of the paper, including discussion of reanalyses, methodology, availability of materials, etc. When detection mechanisms are freely available to use, they can lead to (a surge of) comments when applied by users. Recently, in 2016, one user (the main author of this paper) used `statcheck` software to report potential statistical reporting inconsistencies for 50,000 psychology articles, which led to a large discussion about (automated) online comments \citep{Baker_2016}. The impact of such new possibilities is not to be underestimated, although its potential to constructively contribute to the scientific discussion should also not be \citep{10.1038/540007b}. Nonetheless, the `statcheck` software was well validated prior to this application \citep{Nuijten_2015} and the same principle applies to the application of statistical methods to detect (potential) data fabrication.

Throughout this paper, we inspect statistical methods to detect data fabrication that can be applied to (1) summary results or (2) raw data. 
Even though the available data can look different depending on the structure of a study, there are certain common characteristics of empirical results and the underlying raw data that can be inspected. For example, summary results frequently include means, standard deviations, test-statistics, and $p$-values. Raw data frequently contain at least some variables measured at a interval- or ratio scale \citep{Stevens_1946}. Such common characteristics allow for the development of generic statistical methods that can be applied across a varied set of results.

\section{Theoretical framework}

\subsection{Detecting data fabrication in summary results}

\subsubsection{\textit{P}-values}

The distribution of one $p$-value depends on various factors, but there are specific theoretical boundary conditions \citep{fisher1925}. Based on the population effect size, the precision of the estimate, and the observed effect size, the $p$-value distribution can be uniform (when then null hypothesis is true) or right-skewed (when the alternative hypothesis is true). By extension, any other distribution, after taking into account sampling error, is theoretically not possible. For example, a distribution might show a bump when authors $p$-hack \citep[e.g., Figure 1 in ][]{10.7717/peerj.1935}, but fabricators might create other non-uniform or non-right-skewed distributions, failing to take into account the theoretical boundary conditions.

In order to test whether observed $p$-values transgress this boundary condition, we proposed an adaptation of Fisher's method \citep{10.1186/s41073-016-0012-9}. This adaptation is a simple reversal of the Fisher method \citep{fisher1925}, which was originally introduced as a simple meta-analytic test to indicate presence of an effect; this test is computed as 
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
\label{fishertest}
\end{equation}
where it tests for right-skew (i.e., more smaller $p$-values than larger $p$-values) across the $k$ number of $p$-values. Reversing the Fisher method results in 
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
\label{fisheradjusted}
\end{equation}
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. This threshold is added and can be set to anything. Given the frequent misinterpretation of $p$-values [e.g., as the probability of an effect,  @Goodman2008135;@Altman_1995], researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution and generate only large $p$-values when fabricating nonsignificant results. In that case, it makes  sense to set $t=.05$ or $t=.10$, in order to include only nonsignificant test results.

Two clear exceptions should be taken into account when using this test: (1) model-fit tests and (2) wrongly specified one-tailed tests. When model-fit tests are used, these can quickly result in high $p$-values because of model saturation. For properly specified one-tailed tests, the $p$-value distribution is right-skewed. When wrongly specified, it flips around and becomes left-skew. These two exceptional cases should not be forgotten during the application of this test, but are irrelevant for the rest of this paper due to our study design (see methods of both studies).

\subsubsection{Variances}

Variance or standard deviation estimates are typically reported to indicate spread, but just like the mean there should be sampling error in this estimate (i.e., the variance of the variances) proportionate to the sample size. TheA variance estimate follows a $\chi^2$-distribution, which is dependent on the sample size, that is
\begin{equation}
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
\label{simvar}
\end{equation}
where $N_j$ is the sample size of the $j$th group and $MS_w$ is the normalizing constant resulting in a standardized variance $z_j$. The normalizing constant $MS_w$ is computed as 
\begin{equation}
MS_w=\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}
\label{msw}
\end{equation}
where $s^2_j$ is the observed variance in the $j$th group.

By extension, the expected spread of reported standardized variances creates the benchmark to. This spread can be computed in various ways, such as the standard deviation of the standardized variances (denoted $SD_z$) or as the range of the standardized variances (denoted $max-min_z$).

\subsection{Detecting data fabrication in raw data}

@@

Second, we applied the reversed Fisher method to detect data fabrication to the nonsignificant $p$-values twice: once for the results of gender effects hypothesis in each study and once for the results of the interaction effect hypothesis for each study.
